The Argennon Virtual Machine

### 2021-03-18 tex/avm.tex
Remove htm directory from git and fix readability issues
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

% Preamble
\documentclass[11pt, A4]{report}

% Packages
%\usepackage{amsmath}
\usepackage{graphicx}
\graphicspath{ {../img/} }
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage[ruled,boxed]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{amssymb}
\lstset{
%  basicstyle=\ttfamily,
    mathescape
}
%\binoppenalty=99999
%\relpenalty=99999

@@

% Document
\begin{document}
    \maketitle

@@

    Every block of the Argennon blockchain corresponds to a set of transactions. We store the commitment of this
    transaction set in every block, but we don't keep the set itself. To be able to detect replay attacks, we require
    every signature that a user creates to have a nonce. This nonce consists of the issuance round of the signature
    and a sequence number: \texttt{(issuance,\ sequence)}. When a user creates more than one signature in a round, he
    must sequence his signatures starting from 0 (i.e.~the sequence number restarts from 0 in every round). We define
    a maximum lifetime for signatures, so a signature is invalid if \texttt{currentRound - issuance \textgreater{}\
        maxLifeTime} or if a signature of the same user with a bigger or equal nonce is already used
    (i.e.~is recorded in the blockchain). A nonce is bigger than another nonce if it has an older issuance. If two
    nonces have an equal issuance, the nonce with the bigger sequence number will be considered bigger.

@@

    Every transaction in the Argennon blockchain starts with an \texttt{invoke\_external} instruction which calls a
    special method from ARG smart contract. This method will transfer the proposed fee of the transaction in ARGs
    from a sender account to the fee sink accounts. Argennon has two fee sink accounts: \texttt{execFeeSink} collects
    execution fees and \texttt{dbFeeSink} collects fees for ZK-EDBs. The Protocol decides how to distribute the
    transaction fee between these two fee sink accounts.

@@

    \begin{itemize}
        \item \(G_d\) is an undirected graph.
        \item Every vertex in \(G_d\) corresponds to a transaction and vice versa.
        \item Vertices \(u\) and \(v\) are adjacent in \(G_d\) if and only if \(u\) has a memory location \(L\) in its
        writing list and \(v\) has \(L\) in either its writing list or reading list.
    \end{itemize}

    If we consider a proper vertex coloring of \(G_d\), every color class will give us an independent set of
    transactions which can be executed concurrently. To achieve the highest parallelization, we need to color \(G_d\)
    with minimum number of colors. The \emph{chromatic number} of the memory dependency graph thus shows how good a
    transaction set could be run concurrently.

@@

    After constructing the memory dependency graph of a transaction set, we can use it to construct the
    \emph{execution DAG} of transactions. The execution DAG of a transaction set \(T\) is a directed acyclic
    graph \(G_e\) which has the \emph{execution invariance} property:
    \begin{itemize}
        \item Every vertex in \(G_e\) corresponds to a transaction in \(T\) and vice versa.
        \item Applying the transactions of \(T\) in any order that \emph{respects} \(G_e\) will result in
        the same AVM state.
        \begin{itemize}
            \item An ordering of transactions of \(T\) respects \(G_e\) if for every directed edge \((u,v)\) in \(G_e\)
            the transaction \(u\) comes before the transaction \(v\) in the ordering.
        \end{itemize}
    \end{itemize}

    Having the execution DAG of a set of transactions, using Algorithm~\ref{alg:exec_dag}, we can apply the transaction
    set to the AVM state concurrently, using multiple processor, while we can make sure that the resulted AVM state will
    always be the same no matter how many processor we have used.

    %\includegraphics[width=17cm]{../img/Alg1s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{Ready}{$R_e$}\SetKwData{V}{$v_f$}\SetKwData{Graph}{$G_e$}\SetKwData{Vertices}{$V$}\SetKwData
        {Txns}{$T$}
        \KwData{The execution dag $\Graph = (\Vertices,E)$ of a transaction set \Txns}
        \KwResult{The state of the AVM after applying \Txns with any ordering respecting \Graph}
        \BlankLine
        \Ready $\gets$ the set of all vertices of \Vertices with in degree 0\;
        \While{$\Vertices \neq \varnothing$}
        {
            wait until a new free processor is available\;
            \If{the execution of a transaction was finished}
            {
                remove the vertex of the finished transaction \V from \Graph\;
                \For{each vertex $u \in Adj[\V]$}
                {
                    \If{$u$ has zero in degree}
                    {
                        $\Ready \gets \Ready \cup u$\;
                    }
                }
            }
            \If{$\Ready \neq \varnothing$}
            {
                remove a vertex from \Ready and assign it to a processor\;
            }
        }
        \caption{executing DAG transactions}\label{alg:exec_dag}
    \end{algorithm}

@@

    %\includegraphics[width=17cm]{../img/Alg2s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{Txns}{$T$}\SetKwData{Gd}{$G_d=(V_d,E_d)$}
        \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
        \Input{The memory dependency graph \Gd of a transaction set \Txns\\A proper coloring of $G_d$}
        \Output{An execution dag $G_e=(V_e,E_e)$ for the transaction set \Txns}
        \BlankLine
        $V_e \gets V_d$\;
        $E_e \gets \varnothing$\;
        define a total order on colors of $G_d$\;
        \For{each edge $\{u,v\} \in E_d$}
        {
            \eIf{$color[u] < color[v]$}
            {
                $E_e \gets E_e \cup (u,v)$\;
            }{
                $E_e \gets E_e \cup (v,u)$\;
            }
        }
        \caption{Constructing an execution DAG}\label{alg:gen_dag}
    \end{algorithm}

    The block proposer is responsible for proposing an efficient execution DAG alongside his proposed block which will
    determine the ordering of block transactions and help validators to validate transactions in parallel. Since with
    better parallelization a block can contain more transactions, a proposer is incentivized enough to find a good
    execution DAG for transactions.

@@

    We know that in Argennon every transaction needs to transfer its proposed fee to the \texttt{feeSink} accounts
    first. This essentially makes every transaction a reader and a writer of the memory locations which store the
    balance record of the \texttt{feeSink} accounts. As a result, all transactions in Argennon will be dependant and
    parallelism will be completely impossible. Actually, any account that is highly active, for example the account
    of an exchange or a payment processor, could become a concurrency bottleneck of the system, making all transactions
    which interact with them dependant.

    This problem can be easily solved by using a concurrent counter for storing the balance of this type of
    accounts. A concurrent counter is a data structure which improves concurrency by using multiple memory locations for
    storing a single counter. The value of the concurrent counter is equal to the sum of its sub counters and it can
    be incremented or decremented by incrementing/decrementing any of the sub counters. This way, a concurrent
    counter trades concurrency with memory usage.

    A pseudocode for implementing a concurrent counter which returns an error when the value of the counter
    becomes negative, follows:

    %\includegraphics[width=17cm]{../img/Alg3s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{CC}{Counter}
        \SetKwFunction{Inc}{Increment}\SetKwFunction{Dec}{Decrement}\SetKwFunction{AtomInc}{AtomicIncrement}
        \SetKwFunction{AtomDec}{AtomicDecrement}\SetKwFunction{AtomSet}{AtomicSet}
        \SetKwProg{Fn}{Function}{}{}
        \BlankLine
        \Fn{\Inc{\CC, value, seed}}
        {
            $i \gets seed \bmod \CC.size$\;
           \AtomInc{$\CC.cell[i]$, value}\;
        }
        \BlankLine
        \Fn{\Dec{\CC, value, seed, attempt}}
        {
            \If {attempt = \CC.size}
            {
                restore \CC by adding back the subtracted value\;
                \KwRet{Error}\;
            }
            $i \gets seed \bmod \CC.size$\;
            $i \gets (i + attempt) \bmod \CC.size$\;
            \eIf {$\CC.cell[i] \geq value$}
            {
                \AtomDec{$\CC.cell[i]$, value}\;
            }{
                $r \gets value - \CC.cell[i]$\;
                \AtomSet{$\CC.cell[i]$, $0$}\;
                \Dec{\CC, r, seed, $attempt + 1$}\;
            }
        }
        \caption{Concurrent counter}\label{alg:CC}
    \end{algorithm}

@@

    Concurrent counter data structure is a part of the AVM standard library, and any smart contract can use this data
    structure for storing the balance of highly active accounts.

@@

    In order to further increase the concurrency level of Argennon, we can divide the AVM memory into \emph{chunks}.
    Each memory chunk can be persisted using a different ZK-EDB, hence having its own commitment. Then, the
    consensus on new values of the commitment of any chunk can be achieved by different voting committees.

    If a transaction does not modify a memory chunk and in the transaction ordering of the block it comes after
    any transaction which modifies that chunk, then the execution of that transaction is not needed for calculating
    the new commitment of the chunk. Consequently, the voting committee of the memory chunk can safely ignore such a
    transaction. The execution DAG of transactions can be used for finding and pruning these transactions as
    we see in Algorithm~\ref{alg:prune_dag}.

    %\includegraphics[width=17cm]{../img/Alg4s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{V}{$v_0$}\SetKwData{Graph}{$G_e$}\SetKwData{Chunk}{$C_k$}\SetKwData{Txns}{$T$}
        \SetKwFunction{RPrune}{RecursivePrune}
        \SetKwProg{Fn}{Function}{}{}
        \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
        \Input{An execution dag and a memory chunk \Chunk}
        \Output{The pruned execution dag based on \Chunk}
        \BlankLine
        \For{each vertex \V with out degree $0$}
        {
            \RPrune{\V}\;
        }
        \BlankLine
        \Fn{\RPrune{\V}}
        {
            \If{$Adj[\V] = \varnothing$ {\bf and} \V does not modify \Chunk}
            {
                remove \V from \Graph\;
                \For{each $u$ such that $\V \in Adj[u]$}
                {
                    \RPrune{u}\;
                }
            }
        }
        \caption{Pruning the execution DAG}\label{alg:prune_dag}
    \end{algorithm}

    If we choose chunks in a way that most transactions only modify memory locations of one chunk,
    likely the transactions of a block are divided between voting committees and are validated in parallel.

@@

    To mitigate this problem, for calculating a user's stake at the time step \(t\), instead of using the raw ARG
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ARG balance:
    \[
        S_{u,t} = \min (B_{u,t}, Trust_{u,t})
    \]
    Where:
    \begin{itemize}
        \item \(S_{u,t}\) is the stake of the user \(u\) at the time step \(t\).
        \item \(B_{u,t}\) is the ARG balance of the user \(u\) at the time step \(t\).
        \item \(Trust_{u,t}\) is an estimated trust value for the user \(u\) at the time step \(t\).
    \end{itemize}

    The agreement protocol, at the time step \(t\), will use \(\sum_{u}S_{u,t}\) to determine the required
    number of votes for the confirmation of a block, and we let \(Trust_{u,t} = M_{u,t}\), where \(M_{u,t}\) is the
    exponential moving average of the ARG balance of the user \(u\) at the time step \(t\).

@@

    For calculating the exponential moving average of a user's balance at the time step \(t\), we can use the following
    recursive formula:
    \[
        M_{u,t} = (1 - \alpha) M_{u,t-1} + \alpha B_{u,t} = M_{u,t-1} + \alpha (B_{u,t} - M_{u,t-1})
    \]
    Where the coefficient \(\alpha\) is a constant smoothing factor between \(0\) and \(1\) which represents the
    degree of weighting decrease, A higher \(\alpha\) discounts older observations faster.

>>>>

% Preamble
\documentclass[a4paper]{report}

% Packages
%\usepackage{amsmath}
%\usepackage[a4paper]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage[ruled,boxed]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{amssymb}
\lstset{
%  basicstyle=\ttfamily,
    mathescape
}
%\binoppenalty=99999
%\relpenalty=99999

@@

% Document
\begin{document}
    \maketitle
    \tableofcontents

@@

    Every block of the Argennon blockchain corresponds to a set of transactions. We store the commitment of this
    transaction set in every block, but we don't keep the set itself. To be able to detect replay attacks, we require
    every signature that a user creates to have a nonce. This nonce consists of the issuance round of the signature
    and a sequence number: \texttt{(issuance,\ sequence)}. When a user creates more than one signature in a round, he
    must sequence his signatures starting from 0 (i.e.~the sequence number restarts from 0 in every round). We define
    a maximum lifetime for signatures, so a signature is invalid if \texttt{currentRound - issuance > maxLifeTime} or
    if a signature of the same user with a bigger or equal nonce is already used
    (i.e.~is recorded in the blockchain). A nonce is bigger than another nonce if it has an older issuance. If two
    nonces have an equal issuance, the nonce with the bigger sequence number will be considered bigger.

@@

    Every transaction in the Argennon blockchain starts with an \texttt{invoke\_external} instruction which calls a
    special method from ARG smart contract. This method will transfer the proposed fee of the transaction in ARGs
    from a sender account to the fee sink accounts. Argennon has two fee sink accounts: \texttt{execFeeSink} collects
    execution fees and \texttt{dbFeeSink} collects fees for ZK-EDB servers. The Protocol decides how to distribute the
    transaction fee between these two fee sink accounts.

@@

    \begin{itemize}
        \item \(G_d\) is an undirected graph.
        \item Every vertex in \(G_d\) corresponds to a transaction and vice versa.
        \item Vertices \(u\) and \(v\) are adjacent in \(G_d\) if and only if \(u\) has a memory location \(L\) in its
        writing list and \(v\) has \(L\) in either its writing list or its reading list.
    \end{itemize}

    If we consider a proper vertex coloring of \(G_d\), every color class will give us an independent set of
    transactions which can be executed concurrently. To achieve the highest parallelization, we need to color \(G_d\)
    with minimum number of colors. Thus, the \emph{chromatic number} of the memory dependency graph shows how good a
    transaction set could be run concurrently.

@@

    After constructing the memory dependency graph, we can use it to construct the
    \emph{execution DAG} of transactions. The execution DAG of transaction set \(T\) is a directed acyclic
    graph \(G_e = (V_e,E_e)\) which has the \emph{execution invariance} property:
    \begin{itemize}
        \item Every vertex in \(V_e\) corresponds to a transaction in \(T\) and vice versa.
        \item Executing the transactions of \(T\) in any order that \emph{respects} \(G_e\) will result in
        the same AVM state.
        \begin{itemize}
            \item An ordering of transactions of \(T\) respects \(G_e\) if for every directed edge \((u,v) \in E_e\)
            the transaction \(u\) comes before the transaction \(v\) in the ordering.
        \end{itemize}
    \end{itemize}

    Having the execution DAG of a set of transactions, using Algorithm~\ref{alg:exec_dag}, we can apply the transaction
    set to the AVM state concurrently, using multiple processor, while we can be sure that the resulted AVM state will
    always be the same no matter how many processor we have used.

    %##\includegraphics[width=17cm]{../img/Alg1s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{Ready}{$R_e$}\SetKwData{V}{$v_f$}\SetKwData{Graph}{$G_e$}\SetKwData{Vertices}{$V$}\SetKwData
        {Txns}{$T$}
        \KwData{The execution dag $\Graph = (\Vertices,E)$ of transaction set \Txns}
        \KwResult{The state of the AVM after applying \Txns with any ordering respecting \Graph}
        \BlankLine
        \Ready $\gets$ the set of all vertices of \Vertices with in degree 0\;
        \While{$\Vertices \neq \varnothing$}
        {
            wait until a new free processor is available\;
            \If{the execution of a transaction was finished}
            {
                remove the vertex of the finished transaction \V from \Graph\;
                \For{each vertex $u \in Adj[\V]$}
                {
                    \If{$u$ has zero in degree}
                    {
                        $\Ready \gets \Ready \cup u$\;
                    }
                }
            }
            \If{$\Ready \neq \varnothing$}
            {
                remove a vertex from \Ready and assign it to a processor\;
            }
        }
        \caption{executing DAG transactions}\label{alg:exec_dag}
    \end{algorithm}

@@

    %##\includegraphics[width=17cm]{../img/Alg2s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{Txns}{$T$}\SetKwData{Gd}{$G_d=(V_d,E_d)$}
        \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
        \Input{The memory dependency graph \Gd of transaction set \Txns\\A proper coloring of $G_d$}
        \Output{An execution dag $G_e=(V_e,E_e)$ for the transaction set \Txns}
        \BlankLine
        $V_e \gets V_d$\;
        $E_e \gets \varnothing$\;
        define a total order on colors of $G_d$\;
        \For{each edge $\{u,v\} \in E_d$}
        {
            \eIf{$color[u] < color[v]$}
            {
                $E_e \gets E_e \cup (u,v)$\;
            }{
                $E_e \gets E_e \cup (v,u)$\;
            }
        }
        \caption{Constructing an execution DAG}\label{alg:gen_dag}
    \end{algorithm}

    The block proposer is responsible for proposing an efficient execution DAG alongside his proposed block. This
    execution DAG will determine the ordering of block transactions and help validators to validate transactions in
    parallel. Since with better parallelization a block can contain more transactions, a proposer is incentivized enough
    to find a good execution DAG for transactions.

@@

    We know that in Argennon every transaction needs to transfer its proposed fee to the \texttt{feeSink} accounts
    first. This essentially makes every transaction a reader and a writer of the memory locations which store the
    balance record of the \texttt{feeSink} accounts. As a result, all transactions in Argennon will be dependant and
    parallelism will be completely impossible. Actually, any account that is highly active, for example the account
    of an exchange or a payment processor, could become a concurrency bottleneck in our system which makes all
    transactions interacting with them dependant.

    This problem can be easily solved by using a concurrent counter for storing the balance record of this type of
    accounts. A concurrent counter is a data structure which improves concurrency by using multiple memory locations for
    storing a single counter. The value of the concurrent counter is equal to the sum of its sub counters and it can
    be incremented or decremented by incrementing/decrementing any of the sub counters. This way, a concurrent
    counter trades concurrency with memory usage.

    Algorithm~\ref{alg:CC} implements a concurrent counter which returns an error when the value of the counter
    becomes negative.

    %##\includegraphics[width=17cm]{../img/Alg3s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{CC}{Counter}
        \SetKwFunction{Inc}{Increment}\SetKwFunction{Dec}{Decrement}\SetKwFunction{AtomInc}{AtomicIncrement}
        \SetKwFunction{AtomDec}{AtomicDecrement}\SetKwFunction{AtomSet}{AtomicSet}
        \SetKwProg{Fn}{Function}{}{}
        \BlankLine
        \Fn{\Inc{\CC, value, seed}}
        {
            $i \gets seed \bmod \CC.size$\;
            \AtomInc{$\CC.cell[i]$, value}\;
        }
        \BlankLine
        \Fn{\Dec{\CC, value, seed, attempt}}
        {
            \If {attempt = \CC.size}
            {
                restore \CC by adding back the subtracted value\;
                \KwRet{Error}\;
            }
            $i \gets seed \bmod \CC.size$\;
            $i \gets (i + attempt) \bmod \CC.size$\;
            \eIf {$\CC.cell[i] \geq value$}
            {
                \AtomDec{$\CC.cell[i]$, value}\;
            }{
                $r \gets value - \CC.cell[i]$\;
                \AtomSet{$\CC.cell[i]$, $0$}\;
                \Dec{\CC, r, seed, $attempt + 1$}\;
            }
        }
        \caption{Concurrent counter}\label{alg:CC}
    \end{algorithm}

@@

    Concurrent counter data structure is a part of the AVM standard library, and any smart contract can use this data
    structure for storing the balance record of highly active accounts.

@@

    In order to further increase the concurrency level of Argennon, we can divide the AVM memory into \emph{chunks}.
    Each memory chunk can be persisted using a different ZK-EDB, hence having its own commitment. Then, the
    consensus on new values of the commitment of any chunk can be achieved by a different voting committee.

    If a transaction does not modify a memory chunk and in the transaction ordering of the block it comes after
    any transaction which modifies that chunk, then the execution of that transaction is not needed for calculating
    the new commitment of the chunk. Consequently, the voting committee of that memory chunk can safely ignore such a
    transaction. The execution DAG of transactions can be used for finding and pruning these transactions as
    we see in Algorithm~\ref{alg:prune_dag}.

    %##\includegraphics[width=17cm]{../img/Alg4s.png}
    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{V}{$v_0$}\SetKwData{Graph}{$G_e$}\SetKwData{Chunk}{$C_k$}\SetKwData{Txns}{$T$}
        \SetKwFunction{RPrune}{RecursivePrune}
        \SetKwProg{Fn}{Function}{}{}
        \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
        \Input{An execution dag \Graph and a memory chunk \Chunk}
        \Output{The pruned execution dag based on \Chunk}
        \BlankLine
        \For{each vertex \V with out degree $0$}
        {
            \RPrune{\V}\;
        }
        \BlankLine
        \Fn{\RPrune{\V}}
        {
            \If{$Adj[\V] = \varnothing$ {\bf and} \V does not modify \Chunk}
            {
                remove \V from \Graph\;
                \For{each $u$ such that edge $(u,\V)$ was in \Graph}
                {
                    \RPrune{u}\;
                }
            }
        }
        \caption{Pruning an execution DAG}\label{alg:prune_dag}
    \end{algorithm}

    If we choose chunks in a way that most transactions only modify memory locations of one chunk,
    likely many transactions of a block only need to be validated by one voting committee and can be validated in
    parallel by different committees.

@@

    To mitigate this problem, for calculating a user's stake at time step \(t\), instead of using the raw ARG
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ARG balance:
    \[
        S_{u,t} = \min (B_{u,t}, Trust_{u,t})
    \]
    Where:
    \begin{itemize}
        \item \(S_{u,t}\) is the stake of user \(u\) at time step \(t\).
        \item \(B_{u,t}\) is the ARG balance of user \(u\) at time step \(t\).
        \item \(Trust_{u,t}\) is an estimated trust value for user \(u\) at time step \(t\).
    \end{itemize}

    The agreement protocol, at time step \(t\), will use \(\sum_{u}S_{u,t}\) to determine the required
    number of votes for the confirmation of a block, and we let \(Trust_{u,t} = M_{u,t}\), where \(M_{u,t}\) is the
    exponential moving average of the ARG balance of user \(u\) at time step \(t\).

@@

    For calculating the exponential moving average of a user's balance at time step \(t\), we can use the following
    recursive formula:
    \[
        M_{u,t} = (1 - \alpha) M_{u,t-1} + \alpha B_{u,t} = M_{u,t-1} + \alpha (B_{u,t} - M_{u,t-1})
    \]
    Where the coefficient \(\alpha\) is a constant smoothing factor between \(0\) and \(1\) which represents the
    degree of weighting decrease, A higher \(\alpha\) discounts older observations faster.
